Doublet splitting and linear amplitude characteristics of fusion alpha-particle driven ion cyclotron emission

### 2022-02-06 manuscript.tex
updates post internal review

\begin{abstract}
Ion cyclotron emission (ICE) originating from confined populations of fast ions in toroidal fusion plasmas is an important non-invasive, passive diagnostic for current and next generation devices. The ability to model the ICE signals accurately is an essential step towards inferring the characteristics of confined energetic alpha-particles or fast NBI ions and background plasma. In this paper, the linear growth rates of the magnetoacoustic cyclotron instability, which is the leading explanation for ICE, are calculated in high resolution 2D $(k_\parallel, k_\perp)$ space for parameters corresponding to JET Pulse No. 26148. These calculations shed further light on the origin of doublet-splitting of the cyclotron harmonics as well as the apparent need to invoke nonlinear interactions with 1D3V particle-in-cell studies to account for lower harmonics with substantial amplitudes: the inclusion of signals of ICE from a well resolved range of propagation angles accounts for these effects.
\end{abstract}

@@

The effectiveness of analytical theory to reproduce the observed cyclotron harmonic peaks in the fluctuations of the magnetic field was first explained somewhat by 1D PIC simulations where the onset of nonlinearity appeared to imprint into the signal the amplitudes of each of the harmonics as they are at the end of the linear phase; each mode grows independently until growth is terminated for all modes when the fastest mode enters its nonlinear phase. Analysis shows that only higher cyclotron harmonics are destabilised\cite{McClements2015} to the MCI, however observations clearly indicate power at theoretically stable lower harmonics\cite{Cottrell1988}. In a further step towards accurate recreation of experimental observations, nonlinear phases of hybrid\cite{Carbajal2014} and fully kinetic\cite{Cook2013,Chapman2017} PIC simulations went on to show that wave-wave coupling was responsible for raising the amplitude of lower cyclotron harmonics. Most recently, hybrid 2D simulations have begun to shed light on the 2D3V dynamics of ICE\cite{Carbajal2021}.

To better enable ICE as a diagnostic for confined fast ions, it is necessary to further develop our understanding with a range of tools. In this work, I present the 2D $(k_\parallel, k_\perp)$ growth rates of the MCI, under the conditions of the JET PTE\cite{Cottrell1993} namely Pulse No. 26148 are presented. These growth rates are obtrained from a dispersion relation solver, PlasmaDispersionRelations.jl, written in the Julia programming language\cite{Bezanson2017}. These results narrow the gap between analytical theory and expensive HPC simulations and can be run readily on desktop computers in a matter of minutes, further enabling research into this topic.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are found in
$(k_\parallel, k_\bot)$ space for a ring-beam distribution of energetic minority
alpha-particles and Maxwellian electrons and background thermal majority
deuteron ions. Calculations are made for alpha-particles with ring-beams of two
pitch-angle cosine values: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_1})) \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46m$ and
$\psi_0=55.2^\circ$ are the major radius and angle pitch angle at birth, and
$R_1=4m$ is the major radius of the emission region, see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$u_\parallel=0 u_0$ representing a ring distribution with no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is the speed of an alpha-particle with an energy of 3.6 MeV. The complex frequency solutions, $\omega$, are found corresponding to the fast Alfv{\'e}n branch on a grid of $(k_\parallel, k_\bot)$.

The plasma parameters approximate those found in the region indicated in Fig. 3
of Ref. \cite{Cottrell1993}, which are associated with the continuous trace
signal of Fig. 2 (ibid). The electrons and deuterons, both modelled as
Maxwellians, have temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$, which corresponds a major radius of $4m$.
The ring-beam distribution function, Eq. \ref{eq:alphaf}, is a separable function of $v_\parallel$, i.e. the beam given by Eq. 1, and $v_\perp$, i.e. the ring given by Eq. 2. The component parallel (perpendicular) to the magnetic field is a drifting Maxwellian with thermal speed $v_{\parallel, th}$ ($v_{\perp, th}$) and drift velocity $u_\parallel$ ($u_\perp$). The thermal spread of the alpha particles' distribution function $v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The concentration ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are given by the dispersion relation, Eq. \ref{eq:dettensor}, which is satisfied for sets of $(\omega,\vec{k})$. The hot dielectric tensor\cite{Stix}, $\vec{\vec{\epsilon}}$,  given by Eq. \ref{eq:dielectrictensor} is calculated for each species.

@@

\noindent $J_n$ is the $n$th Bessel function of the first kind and takes the argument $z_s=\frac{k_\perp v_\perp}{\Omega_s}$ where $J_n'=\frac{\partial J_n}{\partial z}$.

The integrals over perpendicular velocity required for calculating the dielectric tensor for Maxwellian distribution functions are given in terms of modified Bessel functions of the first kind for both deuterons and ions i.e. when $u_\perp =0$. The same perpendicular integrals over the alpha-particles' ring distribution is calculated numerically via Gauss-Kronrod\cite{Kronrod1965} quadrature using the GaussGK.jl\cite{QuadGK} library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively, on the value of the integrals is used.

Since all the parallel distribution functions for all species are Maxwellians, the parallel integrals can be calculated from the well known plasma dispersion function\cite{Fried1961}, which is given in the appendix.

The summation over Bessel function indices continues until the relative change of the L2 norm between the current and previous iteration is less than a relative tolerance of $10^{-8}$. Root finding of Eq. \ref{eq:dettensor} is performed by an amalgamation of the Nelder-Mead\cite{Nelder1965} optimisation method with the winding number method, whereby if the Nelder-Mead simplex does not contain a root, as calculated via the winding method, then the optimisation procedure follows the Nelder-Mead method, otherwise if a root is present with the simplex then the simplex bifurcates and the child simplex that contains the root is used for the next iteration. The optimisation iteration procedure finishes when the real extent of the simplex is less than $10^{-4}$ of the alpha-particle cyclotron frequency $\Omega_i$ (the only ion species considered are deuterons and alpha-particles, which have identical cyclotron frequencies) and the imaginary extent is less than $10^{-5}\,\Omega_i$. This approach is found to be adequately robust and faster than the Nelder-Mead method or winding number approach used in isolation of one another. Each 512 by 512 scan in $(k_\parallel, k_\perp)$ takes between 5 and 6 minutes on a laptop.

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data
in the $(\omega,k_\parallel)$ plane where shading instead indicates growth
rates. The simplicity of the consecutive unstable cyclotron harmonics visible in
panel (a) betrays the complexity of the structure in the $(\omega,k_\parallel)$ plane. Integration over $k_\parallel$ projects the solutions onto the frequency axis, which loses the structure and produces narrow harmonics. Fig. \ref{fig:2D_imag_zeropitch} plots shows the same data as Fig. \ref{fig:2D_F12_zeropitch} with damped and normal modes in addition, except shading indicates the growth rates in $(k_\parallel, k_\perp)$ space. Overlaid contours indicate the real frequency in units of ion cyclotron harmonics. Only narrow regions around the cyclotron harmonics are unstable. Overlaid straight traces indicate propagation angle. No Doppler shift is present due to the lack of parallel drift in the alpha-particle ring-beam. The growth rate increases approximately linearly with cyclotron harmonic number, which is consistent with the literature\cite{Dendy1994}. Together these figures serves to introduce a 2D view (growth rates in $(k_\parallel, k_\perp)$ space) of traditionally 1D data (growth rates against frequency). This will become important when a non-zero $u_\parallel$ is taken into account next.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig1.pdf}
    \caption{(colour online) Unstable fast Alfv{\'e}n wave solutions to the
    linear Maxwell-Vlasov system of equations, Eq. \ref{eq:dettensor}, for
    thermal electrons, deuterons and a ring distribution of energetic
    alpha-particles i.e. when there is zero parallel drift, $u_\parallel=0$ in
    Eq. 1. (a) Upper panel: growth rate as a function of frequency where shading
    indicates the parallel wavenumber in units of $\Omega_i/V_A$ of the complex
    frequency solution. Consecutive cyclotron harmonics are increasingly
    unstable and are narrow due to the lack of Doppler shift associated with
    parallel drifts. (b) Lower panel: the same solutions as in panel (a) in the
    $(\omega,k_\parallel)$ plane where shading indicates growth rate. The complex structure of the growth rates that exists in wavenumber space is projected into sharp harmonics on the frequency axis.}
    \label{fig:2D_F12_zeropitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig2.pdf}
    \caption{(colour online) Shading indicates the growth rate of fast Alfv{\'e}n waves excited by a ring of fusion born alpha-particles, as a function of parallel wavenumber on the ordinate and perpendicular wavenumber on the abscissa. No solution is found where shading is absent. The alpha-particle distribution function is given by Eq. \ref{eq:alphaf} with $u_{\parallel} = 0$. Consecutive straight traces indicate the propagation angle with respect to the magnetic field: the horizontal line is exactly perpendicular; solid traces fan out at intervals of five degrees; and dashed traces indicate intervals of one degree. Curved grey traces indicate the location in $(k_\parallel, k_\perp)$ of the cyclotron harmonics (i.e. the real component of the complex $\omega$ solutions), which are annotated along the left hand side and upper edge of the figure. Frequencies are in units of the alpha-particle cyclotron frequency, and wavenumbers are in alpha-particle cyclotron frequencies per Alfv{\'e}n speed. Note the up-down symmetry due to the lack of parallel drift in the energetic alpha particles.}
    \label{fig:2D_imag_zeropitch}
\end{figure}

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physical pitch-angle
case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch}. There are several points worth noting. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $11^{th}$ harmonic is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Dendy et al 1994 (also
Fig. 9 Dendy et al 1995) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. The doublet splitting arises due to the dip in growth rate where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign. The effect is
exaggerated for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation; see e.g. the $12^{th}$ harmonic in Fig.
\ref{fig:2D_imag_physicalpitch} near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $\Omega_{i}$ is the ion cyclotron
frequency and $V_A$ is the Alfv{\'e}n speed. Other possible mechanisms for
doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998}. Qualitatively, Fig. \ref{fig:2D_F12_physicalpitch}(a) resembles Fig. 2 of Cottrell 1993 in that the signal below $\sim 7$ cyclotron harmonics is lower and flatter with well defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change larger and the harmonics are less easily identifiable. These features can only be recreated with a 2D calculation for these distributions functions. Note that Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is characterised by narrow regions of instability near Doppler shifted cyclotron harmonics that are stabilised when $k_\parallel=0$ but unstable at larger negative parallel wavenumbers. The higher frequency behaviour in the region $\omega\geq 7\Omega_i$ is dominated by broader regions of instability in frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have a complex structure in wavenumber space. Modes propagating exactly perpendicular become stable in the region $k_\perp > 13 \Omega_i/V_A$ whilst obliquely propagating modes remain unstable; see Fig. 9 of Dendy et al 1995, which plots the analytic expression for the growth rate of the magnetoacoustic cylcotron instability, i.e. the instability thought responsible for ICE, at 5 slices of propagation angle. The Doppler shift of the growing modes can reach $\approx \Omega_i$ for large propagation angles at lower frequencies and also for moderately oblique angles at higher frequencies; see, for example, the mode associated with the $15^{th}$ cyclotron harmonic at, $(k_\parallel \simeq -\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation angle of $93^\circ$). Furthermore, the breadth of the harmonics can be attributed to the contributions from multiple propagation angles even though the lower harmonics are stable for forward travelling and perpendicular propagation angles.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig3.pdf}
    \caption{(colour online) The imaginary component of the frequencies for the
    case with parallel drift, $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$.
    Panel (a) plots growth rate against frequency where shading indicates
    parallel wavenumber in units of $\Omega_i/V_A$, and (b) plots the solutions
    in the $(\omega,k_\parallel)$ plane where shading indicates growth rate. Note
    that the Doppler shift associated with the non-zero parallel drift speed
    broadens the unstable peaks in frequency. The $10^{th}$ and $11^{th}$ cyclotron
    harmonics are split into doublets, i.e. they have two peaks because both
    the growth rate and real frequency vary with wavenumber. Compare with
    Fig. 2 of Cottrel et al 1993.}
    \label{fig:2D_F12_physicalpitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig4.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except for a ring-beam of alpha particles, with a physical pitch-angle value resulting in $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$. The non-zero parallel drift speed causes up-down asymmetry; this figure for a pitch-angle cosine of the opposite sign, $u_{\parallel} = 0.646 u_0$, is identical expect flipped about $k_\parallel=0$.}
    \label{fig:2D_imag_physicalpitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig5.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except that the shading indicates the real part of the smoothed solution normalised by the high frequency expression for the fast Alfv{\'e}n wave set out by Eq. 25 of Ref. \cite{Dendy1994}, which is also Eq. 11 in Ref. \cite{McClements1996}. The differences in real frequency for the two sets of calculations, physical pitch-angle cosine of $-0.646$ and zero, are negligible.}
    \label{fig:2D_real_zeropitch}
\end{figure}

@@

This paper presents the linear Maxwell-Vlasov dispersion relation of a plasma consisting of thermal electrons, thermal deuteron fuel ions and a minority energetic ring-beam of fusion born alpha-particles. These high resolution $(k_\parallel, k_\perp)$ calculations revisit ICE observations from JET Pulse No. 26148\cite{Cottrell1993,Dendy1995} and reproduce several important features: the step change in signal strength from $\omega < 6$ to $\omega > 6$; the doublet splitting of peaks at $\omega \approx 10\Omega_i$ on Fig. \ref{fig:2D_F12_physicalpitch}, with origins visible in the fine $(k_\parallel, k_\perp)$ structure in Fig. \ref{fig:2D_imag_physicalpitch}. The Doppler shift for large $k_\parallel$ becomes of the order of the alpha-particle cyclotron frequency, which makes it difficult to determine which parts of the signal come from which cyclotron harmonic numbers. Further effort is required to account for the extra doublets visible in the continuous trace associated with JET Pulse No. 26148 in Fig. 2 of Ref. \cite{Cottrell1993}.

>>>>

\begin{abstract}
Ion cyclotron emission (ICE) originating from confined populations of fast ions
  in toroidal fusion plasmas is an important non-invasive, passive diagnostic
  for current and next generation devices. The ability to model the ICE signals
  accurately is an essential step towards inferring the characteristics of
  confined energetic alpha-particles or fast NBI ions and background plasma. In
  this paper, the linear growth rates of the magnetoacoustic cyclotron
  instability, which is the leading explanation for ICE, are calculated in high
  resolution 2D $(k_\perp, k_\parallel)$ space for parameters corresponding to
  JET Pulse No. 26148. These calculations shed further light on the origin of
  doublet-splitting of the cyclotron harmonics as well as the apparent need to
  invoke nonlinear interactions with 1D3V particle-in-cell studies to account
  for lower harmonics with substantial amplitudes: the inclusion of signals of
  ICE from a well resolved range of propagation angles can also account for
  these effects.
\end{abstract}

@@

The effectiveness of analytical theory to reproduce the observed cyclotron
harmonic peaks in the fluctuations of the magnetic field was first explained
somewhat by 1D PIC simulations. In these simulations the onset of nonlinearity
appeared to imprint the amplitudes of each of the harmonics into the signal as
they are at the end of the linear phase; each mode grows independently until 
growth is terminated for all modes when the fastest mode enters its nonlinear
phase meaning that both the amount of free energy available for instability has
decreased and that the unstable energy ion distribution function has been
perturbed into a more stable state. Analysis shows that only higher cyclotron
harmonics are destabilised\cite{McClements2015} to the MCI, however observations
clearly indicate power at theoretically stable lower
harmonics\cite{Cottrell1988}. In a further step towards accurate recreation of
experimental observations, nonlinear phases of hybrid\cite{Carbajal2014} and
fully kinetic\cite{Cook2013,Chapman2017} PIC simulations went on to show that
wave-wave coupling was responsible for raising the amplitude of lower cyclotron
harmonics. Most recently, hybrid 2D simulations have begun to shed light on the
2D3V dynamics of ICE\cite{Carbajal2021}.

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. In this work, I present
the 2D $(k_\perp, k_\parallel)$ growth rates of the MCI under the conditions of
outer mid-plane of the JET PTE\cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
PlasmaDispersionRelations.jl, written in the Julia programming
language\cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run readily on desktop computers
in a matter of minutes, further enabling research into this topic. Note that
this study focusses on edge-ICE, named for the region from which the signal
arises, as opposed to core-ICE\cite{Ochoukov2018}.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are found in
$(k_\parallel, k_\bot)$ space for a ring-beam distribution of energetic minority
alpha-particles and Maxwellian electrons and background thermal majority
deuteron ions. Calculations are made for alpha-particles with ring-beams of two
values of pitch-angle cosine: $u_\parallel=-0.646 u_0 \approx
-\cos(\arcsin(\sqrt{R_0 \sin(\psi_0)/R_{ICE}})) \simeq -1.1 V_A$, which is
associated with the ion-cyclotron emission observed on JET Pulse No. 26148 for
which $R_0=3.46$ m and
$\psi_0=55.2^\circ$ are the major radius and pitch angle at birth, and
$R_{ICE}=4$ m is the major radius of the emission region, see
Figs. 3 and 14 and associated text in Ref. \cite{Cottrell1993}; and
$u_\parallel=0 u_0$ representing a ring distribution with no parallel drift, see
Fig. 1(a) of Ref \cite{Cook2013}. Here, $V_A$ is the Alfv{\'en} speed and $u_0=\sqrt{u_\parallel^2 + u_\perp^2}$ is
the speed of an alpha-particle with an energy of 3.6 MeV. The complex frequency
solutions, $\omega$, along the fast Alfv{\'e}n branch are found on a grid of $(k_\parallel, k_\bot)$.

The plasma parameters approximate those found in the region indicated in Fig. 3
of Ref. \cite{Cottrell1993}, which are associated with the continuous trace
signal of Fig. 2 (ibid). The electrons and deuterons have temperatures of 1 keV and the electrons have a number density
of $1.7\times 10^{19}$ $\mathrm{m^{-3}}$. The magnetic field is $2.07$
$\mathrm{T}$, which corresponds a major radius of $4$ m.
The ring-beam distribution function, Eq. \ref{eq:alphaf}, is a separable
function of the beam in $v_\parallel$ given by Eq. 1 and the ring in $v_\perp$
given by Eq. 2. This distribution function is parameterised by the drift 
velocity parallel to the magnetic field $u_\parallel$, the drift speed 
perpendicular to the magnetic
field $u_{\perp}$, and the thermal velocity in both directions $v_{th}$. The
thermal spread of the alpha particles' distribution function 
$v_{th}=v_{\parallel,th}=v_{\perp,th}=u_0/100$. The number density
ratio of alpha-particles to deuterons is $1.5\times10^{-4}$.

@@

Solutions to the linearised Maxwell-Vlasov system of equations are given by the
dispersion relation, Eq. \ref{eq:dettensor}, which is satisfied for sets of
$(\omega,\vec{k})$ given the distribution functions of all the species present.
The hot dielectric tensor\cite{Stix}, $\vec{\vec{\epsilon}}$, Eq. \ref{eq:dielectrictensor} is calculated for each species.

@@

\noindent $J_n$ is the $n$th Bessel function of the first kind taking the
argument $z_s=k_\perp v_\perp / \Omega_s$ and $J_n'=\frac{\partial J_n}{\partial z}$.

The integrals over perpendicular velocity required for calculating the
dielectric tensor for Maxwellian distribution functions are given in terms of
modified Bessel functions of the first kind\cite{Stix} for both electrons and
ldeuterons i.e. when $u_\perp =0$. The same perpendicular integrals over the
alpha-particles' ring distribution is calculated numerically via
Gauss-Kronrod\cite{Kronrod1965} quadrature using the GaussGK.jl\cite{QuadGK}
library. An absolute and relative tolerance of $0$ and $10^{-8}$, respectively,
on the value of the integrals is used.

Since all the parallel distribution functions for all species are Maxwellians,
the parallel integrals can be calculated from the well known plasma dispersion
function\cite{Fried1961}, which is laid out in the appendix.

The summation over Bessel function indices in Eq. \ref{eq:dielectrictensor}
continues until the relative change of the L2 norm between the current and
previous iteration is less than a tolerance of $10^{-8}$. Root finding of Eq.
\ref{eq:dettensor} is performed by an amalgamation of the
Nelder-Mead\cite{Nelder1965} optimisation method with the winding number method,
whereby if the Nelder-Mead simplex does not contain a root, as calculated via
the winding method, then the optimisation procedure follows the Nelder-Mead
method, otherwise a non-zero winding number indicates that a root is present
within the simplex and the simplex
bifurcates. The child simplex that contains the root is used for the next
iteration. The optimisation iteration procedure finishes when the real extent of
the simplex is less than $10^{-4}\, \Omega_i$ and the imaginary extent is less
than $10^{-5}\,\Omega_i$, where $\Omega_i$ is the cyclotron frequency of both
the alpha-particles and deuterons since they share a charge-mass ratio.
This approach is found to be adequately robust and
faster than the Nelder-Mead method or winding number approach used in isolation
of one another. Each 512 by 512 scan in $(k_\perp, k_\parallel)$ takes between 5
and 6 minutes on a laptop.

@@

The growth rates of unstable modes for the case where the pitch-angle cosine
value is zero ($u_\parallel = 0$ in Eq. 1) are plotted in Fig.
\ref{fig:2D_F12_zeropitch}(a) against the associated real frequency (only modes
with growth rates in excess of the solver tolerance are shown i.e. $\Im(\omega)
> 10^{-5} \Omega_i$). In this panel shading indicates parallel wavenumber in
units of $\Omega_i/V_A$. Fig. \ref{fig:2D_F12_zeropitch}(b) plots the same data
in the $(\omega,k_\parallel)$ plane where shading instead indicates growth
rates. The simplicity of the consecutive unstable cyclotron harmonics visible in
panel (a) betrays the complexity of the structure in the $(\omega,k_\parallel)$
plane. Integration over $k_\parallel$ projects the solutions onto the frequency
axis, which loses the structure and produces narrow harmonics. Figure
\ref{fig:2D_imag_zeropitch} shows the same data as Fig.
\ref{fig:2D_F12_zeropitch} in $(k_\perp, k_\parallel)$ space with damped and
normal modes in addition, except shading indicates the growth rates. Overlaid
contours indicate the real frequency in units of ion cyclotron harmonics. Only
narrow regions around the cyclotron harmonics are unstable. Overlaid straight
traces indicate propagation angle. No Doppler shift is present due to the lack
of parallel drift in the alpha-particle ring-beam. The growth rate increases
approximately linearly with cyclotron harmonic number, which is consistent with
the literature\cite{Dendy1994}. Together these figures serve to introduce a 2D
view (growth rates in $(k_\perp, k_\parallel)$ space) of traditionally 1D data
(growth rates against frequency). This will become important when a non-zero
$u_\parallel$ is taken into account next.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig1.pdf}
    \caption{(colour online) Unstable fast Alfv{\'e}n wave solutions to the
    linear Maxwell-Vlasov system of equations, Eq. \ref{eq:dettensor}, for
    thermal electrons, deuterons and a ring distribution of energetic
    alpha-particles i.e. when there is zero parallel drift, $u_\parallel=0$ in
    Eq. 1. (a) Upper panel: growth rate as a function of real frequency where shading
    indicates the parallel wavenumber, $k_\parallel$, in units of $\Omega_i/V_A$
    of the complex frequency solution. Consecutive cyclotron harmonics are
    increasingly unstable and are narrow due to the lack of Doppler shift 
    associated with parallel drifts. (b) Lower panel: the same solutions
    as in panel (a) in the $(\omega,k_\parallel)$ plane where shading indicates 
    growth rate. The complex structure of the growth rates that exists in
    wavenumber space is projected into sharp harmonics on the frequency axis.}
    \label{fig:2D_F12_zeropitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig2.pdf}
    \caption{(colour online) Shading indicates the growth rate of fast
    Alfv{\'e}n waves excited by a ring of fusion born alpha-particles, as a
    function of parallel wavenumber on the ordinate and perpendicular wavenumber
    on the abscissa. No solution is found where shading is absent. The
    alpha-particle distribution function is given by Eq. \ref{eq:alphaf} with
    $u_{\parallel} = 0$. Consecutive straight traces indicate the propagation
    angle with respect to the magnetic field: the horizontal line is exactly
    perpendicular; solid traces fan out at intervals of five degrees; and dashed
    traces indicate intervals of one degree. Curved grey traces indicate the
    location in $(k_\perp, k_\parallel)$ of the cyclotron harmonics (i.e. the
    real component of the complex $\omega$ solutions), which are annotated along
    the left hand side and upper edge of the figure. Frequencies are in units of
    the alpha-particle cyclotron frequency, and wavenumbers are in
    alpha-particle cyclotron frequencies per Alfv{\'e}n speed. Note the up-down
    symmetry due to the lack of parallel drift in the energetic alpha
    particles.}
    \label{fig:2D_imag_zeropitch}
\end{figure}

Fig. \ref{fig:2D_F12_physicalpitch} plots solutions for the physically more
realistic pitch-angle
case ($u_\parallel = -0.646 u_0$) in the same manner as Fig.
\ref{fig:2D_F12_zeropitch}. There are several points worth noting. First, the
growth rate is no longer linear with cyclotron harmonic number. Second,
harmonics are broader due to the non-zero parallel drift that gives rise to
strong Doppler shift with increasing $|k_\parallel|$. Third, doublet splitting
of the $10^{th}$ and $11^{th}$ harmonics is observed: see Fig. 2 of Cottrell 1993 for a single
signal showing multiple doublet splittings; Fig. 3 of Dendy et al 1994 (also
Fig. 9 Dendy et al 1995) shows slices at constant propagation angle that when
all superimposed point towards doublet splitting, albeit under-resolved in
wavenumber space. The following describes how these three features are
connected. Doublet splitting arises due to the dip in growth rate along the 
band of instability where
$k_\parallel$ goes through zero combined with the change in sign of the Doppler
shift of the real frequency as $k_\parallel$ changes sign. The effect is
larger for higher harmonics where the growth rate diminishes to zero for
perpendicular propagation; see e.g. the $12^{th}$ harmonic in Fig.
\ref{fig:2D_imag_physicalpitch} near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed. Other possible mechanisms for
doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998}. Qualitatively, Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles Fig. 2 of Cottrell 1993 in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stabilised when $k_\parallel=0$ but unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable; see Fig. 9 of Dendy et al 1995, which plots
the analytic expression for the growth rate of the MCI at 5 slices of
propagation angle. The Doppler shift of the growing modes can reach $\approx
\Omega_i$ for large propagation angles at lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and perpendicular
propagation angles.

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig3.pdf}
    \caption{(colour online) The imaginary component of the frequencies for the
    case with parallel drift $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$.
    Panel (a) plots growth rate against frequency where shading indicates
    parallel wavenumber in units of $\Omega_i/V_A$, and (b) plots the solutions
    in the $(\omega,k_\parallel)$ plane where shading indicates growth rate. Note
    that the Doppler shift associated with the non-zero parallel drift speed
    broadens the unstable peaks in frequency. The $10^{th}$ and $11^{th}$ cyclotron
    harmonics are split into doublets, i.e. they have two peaks. Compare with
    Fig. 2 of Cottrel et al 1993.}
    \label{fig:2D_F12_physicalpitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig4.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except for
    a ring-beam of alpha particles with a more realistic pitch-angle value
    $u_{\parallel} = -0.646 u_0 \simeq -1.1 V_A$. The non-zero parallel drift
    speed causes up-down asymmetry; this figure for a pitch-angle cosine of the
    opposite sign, $u_{\parallel} = 0.646 u_0$, is identical expect flipped
    about $k_\parallel=0$.}
    \label{fig:2D_imag_physicalpitch}
\end{figure}

\begin{figure}[ht!]
    % \centering
    \raggedleft
\includegraphics[scale=0.9]{fig5.pdf}
    \caption{(colour online) As in Fig. \ref{fig:2D_imag_zeropitch}, except that
    the shading indicates the real part of the smoothed solution normalised by
    the high frequency expression for the fast Alfv{\'e}n wave set out in Eq. 25
    of Ref. \cite{Dendy1994}, which is also Eq. 11 in Ref.
    \cite{McClements1996}. The differences in real frequency for the two sets of
    calculations, a pitch-angle cosine of $-0.646$ and zero, are
    negligible.}
    \label{fig:2D_real_zeropitch}
\end{figure}

@@

This paper presents the linear Maxwell-Vlasov dispersion relation of a plasma
consisting of thermal electrons, thermal deuteron fuel ions and a minority
energetic ring-beam of fusion born alpha-particles. These high resolution
$(k_\perp, k_\parallel)$ calculations revisit ICE observations from JET Pulse
No. 26148\cite{Cottrell1993,Dendy1995} and reproduce several important features:
the step change in signal strength from $\omega < 6$ to $\omega > 6$; the
doublet splitting of peaks at $\omega \approx 10\Omega_i$ in Fig.
\ref{fig:2D_F12_physicalpitch}, with origins visible in the fine $(k_\parallel,
k_\perp)$ structure in Fig. \ref{fig:2D_imag_physicalpitch}. The Doppler shift
for large $k_\parallel$ becomes of the order of the alpha-particle cyclotron
frequency, which makes it difficult to determine which parts of the signal come
from which cyclotron harmonic numbers. Further effort is required to account for
the extra doublets visible in the continuous trace associated with JET Pulse No.
26148 in Fig. 2 of Ref. \cite{Cottrell1993}.
